### 2023-06-05 paper/localpsf.tex
editing pass on paper with Tucker. Paper is ready for Noemi

%Existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, which we cannot do. 
We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applies. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of $\phi_x$\todo{the width of the support?} is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Our moment method estimates the support of $\phi_x$ more accurately than random noise probing, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
We form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of $\Aop$\todo{here a reviewer might complain that $\Aop$ is an infinite dimensional operator and H-matrix methods are only defined for finite dimensional operators}. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.

@@

We modify $\mathbf{A}_H$ to make it symmetric and remove negative eigenvalues. First, we symmetrize $\mathbf{A}_H$ via
$
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right).
$
Next, we find negative eigenvalues and their corresponding eigenvectors for the generalized eigenvalue problem
$\mathbf{A}_H^{\text{sym}} \mathbf{u} = \lambda \mathbf{R}_H$
using a Cayley shift-and-invert Krylov scheme \cite{LehoucqEtAl98}. We flip the signs of these eigenvalues to be positive instead of negative (i.e., $\lambda \rightarrow |\lambda|$) by performing a low rank update to $\mathbf{A}_H^{\text{sym}}$. The primary computational task in the Cayley shift-and-invert scheme is the solution of shifted linear systems of the form
$\left(\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H\right) x = b,$
for a small number of positive shifts $\mu_i$. We solve these linear systems by factorizing the matrices $\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H$ using fast H-matrix methods. We compute and flip all eigenvalues $\lambda < \tau$ which are less than some threshold $\tau \in (-1,0]$\todo{notation conflict, $\tau$ used as threshold here and previously as a number of standard deviations in choosing batches of points from Gaussian info of impulse responses}. By choosing $\tau > -1$, we ensure that the modified version of $\mathbf{A}_H^{\text{sym}} + \mathbf{R}_H$ is positive definite. Choosing $\tau=0$ would remove all erroneous negative eigenvalues. However, this is computationally infeasible if $\Aop$ has a large or infinite cluster of eigenvalues near zero, a common situation for Hessians in ill-posed inverse problems. We therefore recommend choosing $\tau < 0$, and we use $\tau=-0.1$ in our numerical experiments.

@@

The computational cost of our method may be divided into the costs to perform the following tasks: (1) Computing impulse response moments and batches; (2) Building the H-matrix; (3) Performing linear algebra operations with the H-matrix. This may optionally include a symmetric positive semi-definite modification of the H-matrix. In target applications, (1) is the dominant cost because applying $\Aop$ to a vector requires an expensive computational procedure such as solving a PDE, and (1) is the only step that requires applying $\Aop$ to vectors.
%In practice, (1) is the dominant cost, because (1) is the only task that requires applying $\Aop$ to vectors, and  
All operations that do not require applications of $\Aop$ to vectors are nearly linear, and therefore scalable, in the size of the problem, $N$.
We now describe these costs in detail. For convenience, Table~\ref{tab:vars} lists variable symbols and their approximate sizes.

@@

We use our method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and transport
and flow of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). 
To reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose cost functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using Morozov's discrepancy
principle~\cite{Vogel02}. 
%The optimal value of $\gamma$ is used for
%  $\gamma =  7.3 \times 10^3$ for Stokes  Figure~\ref{fig:gamma_sweep}
%all the numerical results except for the studies where we study how
%changing $\gamma$ impacts the effectiveness of the proposed
%preconditioner.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit). The 
%number of impulse response batches used to build these 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. H-matrix approximations of the Gauss-Newton Hessian, 
$$\bm{H}_\text{gn}\approx \widetilde{\bm{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\bm{H}}$ to be symmetric positive definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel. Our results show that~$\widetilde{\bm{H}}$ is a good preconditioner for both~$\bm{H}_\text{gn}$ and~$\bm{H}$. 
%The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%

@@

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure \ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization determined by the Morozov discrepancy principle for each noise level. The true basal sliding
coefficient field and the true velocity field, which is obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$.}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$.}
		\label{fig:stokes_velocity}
	\end{subfigure}
    \end{center}
	\caption{
		(\ref{fig:ice_mountain_mesh}) Bird's eye view of the ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(\ref{fig:true_beta}) True parameter, $\basalfriction_\text{true}$. (\ref{fig:stokes_velocity}) True velocity $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

	\caption{The log basal sliding parameter, with colorscale as in Figure~\ref{fig:true_beta}, computed from the PDE-constrained optimization problem with noise levels: $25\%$ (left), $5.0\%$ (middle), and $1.0\%$ (right).}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ \\
				0 &
				1 & 4 & 1.9e+7 &
				3 & 8 & 1.9e+7 &
				1 & 4 & 1.9e+7 \\
				1 &
				2 & 6  & 6.1e+6 &
				8 & 18 & 8.4e+6 &
				2 & 6  & 6.1e+6 \\
				2 &
				4 & 10 & 2.6e+6 &
				16 & 34 & 4.1e+6 &
				4 & 10 & 2.6e+6 \\
				3 &
				2 & 6+22 & 6.9e+5 &
				34 & 70 & 1.8e+6 &
				14 & 30 & 6.9e+5 \\
				4 &
				3 & 8 & 4.4e+4 &
				52 & 106 & 5.6e+5 &
				29 & 60 & 1.3e+5 \\
				5 &
				5 & 12 & 2.2e+3 &
				79 & 160 & 9.4e+4 &
				38 & 78 & 1.0e+4 \\
				6 &
				0 & 2 & 1.1e+1 &
				102 & 206 & 6.5e+3 &
				58 & 118 & 1.8e+2 \\
				7 &
				--- & --- & --- &
				151 & 304 & 1.2e+2 &
				0 & 2 & 5.5e-1 \\
				8 & 
				--- & --- & --- &
				0 & 2 & 2.9e-1 &
				--- & --- & --- \\
				\hline
				Total & 
				17 & 70 & --- &
				445 & 908 & --- &
				146 & 308 & --- \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{Convergence history for solving the Stokes inverse problem using inexact Newton PCG. The Newton iteration is terminated when $||\mathbf{g}_k|| < 10^{-6} ||\mathbf{g}_0||$, where $\mathbf{g}_i$ is the gradient at the $i$th iteration. Preconditioners shown are our method with five batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}_k$. Columns titled $\|\bm{g}\|$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}_k$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the line search going from $\bm{\basalfriction}_{k-1}$ to $\bm{\basalfriction}_k$, plus one adjoint Stokes solve to compute the gradient at $\bm{\basalfriction}_k$, plus one incremental forward and one incremental adjoint Stokes solve per PCG iteration for solving the Newton system. Under PSF (5) and in row Iter 3, we write $6+22$ to indicate that $6$ Stokes solves were used during the standard course of the iteration, and $22$ Stokes solves were used to build the PSF (5) preconditioner.}
	\label{tab:newton_convergence_table}
\end{table}

>>>>

%Existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, which we cannot do. 
We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applies. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of the support $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Our moment method estimates the support of $\phi_x$ more accurately than random noise probing, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
We form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.

@@

We modify $\mathbf{A}_H$ to make it symmetric and remove negative eigenvalues via the following procedure. First, we symmetrize $\mathbf{A}_H$ via
$
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right).
$
Next, we find negative eigenvalues and their corresponding eigenvectors for the generalized eigenvalue problem
$\mathbf{A}_H^{\text{sym}} \mathbf{u} = \lambda \mathbf{R}_H$
using a Cayley shift-and-invert Krylov scheme \cite{LehoucqEtAl98}. We flip the signs of these eigenvalues to be positive instead of negative (i.e., $\lambda \rightarrow |\lambda|$) by performing a low rank update to $\mathbf{A}_H^{\text{sym}}$. The primary computational task in the Cayley shift-and-invert scheme is the solution of shifted linear systems of the form
$\left(\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H\right) x = b,$
for a small number of positive shifts $\mu_i$. We solve these linear systems by factorizing the matrices $\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H$ using fast H-matrix methods. We compute and flip all eigenvalues $\lambda < \epsilon_\text{flip}$ which are less than some threshold $\epsilon_\text{flip} \in (-1,0]$. By choosing $\epsilon_\text{flip} > -1$, we ensure that the modified version of $\mathbf{A}_H^{\text{sym}} + \mathbf{R}_H$ is positive definite. Choosing $\epsilon_\text{flip}=0$ would remove all erroneous negative eigenvalues. However, this is computationally infeasible if $\Aop$ has a large or infinite cluster of eigenvalues near zero, a common situation for Hessians in ill-posed inverse problems. We therefore recommend choosing $\epsilon_\text{flip} < 0$, and we use $\epsilon_\text{flip}=-0.1$ in our numerical experiments.

@@

The computational cost of our method may be divided into the costs to perform the following tasks: (1) Computing impulse response moments and batches; (2) Building the H-matrix; (3) Performing linear algebra operations with the H-matrix. This may optionally include the symmetric positive semi-definite modifications described in Section \ref{sec:flip_eigs}. In target applications, (1) is the dominant cost because applying $\Aop$ to a vector requires an expensive computational procedure such as solving a PDE, and (1) is the only step that requires applying $\Aop$ to vectors.
%In practice, (1) is the dominant cost, because (1) is the only task that requires applying $\Aop$ to vectors, and  
All operations that do not require applications of $\Aop$ to vectors are nearly linear, and therefore scalable, in the size of the problem, $N$.
We now describe these costs in detail. For convenience, Table~\ref{tab:vars} lists variable symbols and their approximate sizes.

@@

We use our method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and transport
and flow of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). 
To reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose cost functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using the Morozov discrepancy
principle~\cite{Vogel02}. 
%The optimal value of $\gamma$ is used for
%  $\gamma =  7.3 \times 10^3$ for Stokes  Figure~\ref{fig:gamma_sweep}
%all the numerical results except for the studies where we study how
%changing $\gamma$ impacts the effectiveness of the proposed
%preconditioner.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit). The 
%number of impulse response batches used to build these 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. H-matrix approximations of the Gauss-Newton Hessian, 
$$\bm{H}_\text{gn}\approx \widetilde{\bm{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\bm{H}}$ to be (approximately) symmetric positive semi-definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel. Our results show that~$\widetilde{\bm{H}}$ is a good preconditioner for both~$\bm{H}_\text{gn}$ and~$\bm{H}$. 
%The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%

@@

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure~\ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization is determined by the Morozov discrepancy principle for each noise level. The true basal sliding
coefficient field and the true velocity field, which is obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$.}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$.}
		\label{fig:stokes_velocity}
	\end{subfigure}
    \end{center}
	\caption{\textbf{(Ice sheet)} 
		(\ref{fig:ice_mountain_mesh}) Bird's eye view of the ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(\ref{fig:true_beta}) True parameter, $\basalfriction_\text{true}$. (\ref{fig:stokes_velocity}) True velocity $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

	\caption{\textbf{(Ice sheet)} The log basal sliding parameter, with colorscale as in Figure~\ref{fig:true_beta}, computed from the PDE-constrained optimization problem with noise levels: $25\%$ (left), $5.0\%$ (middle), and $1.0\%$ (right).}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ \\
				0 &
				1 & 4 & 1.9e+7 &
				3 & 8 & 1.9e+7 &
				1 & 4 & 1.9e+7 \\
				1 &
				2 & 6  & 6.1e+6 &
				8 & 18 & 8.4e+6 &
				2 & 6  & 6.1e+6 \\
				2 &
				4 & 10 & 2.6e+6 &
				16 & 34 & 4.1e+6 &
				4 & 10 & 2.6e+6 \\
				3 &
				2 & 6+22 & 6.9e+5 &
				34 & 70 & 1.8e+6 &
				14 & 30 & 6.9e+5 \\
				4 &
				3 & 8 & 4.4e+4 &
				52 & 106 & 5.6e+5 &
				29 & 60 & 1.3e+5 \\
				5 &
				5 & 12 & 2.2e+3 &
				79 & 160 & 9.4e+4 &
				38 & 78 & 1.0e+4 \\
				6 &
				0 & 2 & 1.1e+1 &
				102 & 206 & 6.5e+3 &
				58 & 118 & 1.8e+2 \\
				7 &
				--- & --- & --- &
				151 & 304 & 1.2e+2 &
				0 & 2 & 5.5e-1 \\
				8 & 
				--- & --- & --- &
				0 & 2 & 2.9e-1 &
				--- & --- & --- \\
				\hline
				Total & 
				17 & 70 & --- &
				445 & 908 & --- &
				146 & 308 & --- \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{\textbf{(Ice sheet)} Convergence history for solving the Stokes inverse problem using inexact Newton PCG. The Newton iteration is terminated when $||\mathbf{g}_k|| < 10^{-6} ||\mathbf{g}_0||$, where $\mathbf{g}_i$ is the gradient at the $i$th iteration. Preconditioners shown are our method with five batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}_k$. Columns titled $\|\bm{g}\|$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}_k$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the line search going from $\bm{\basalfriction}_{k-1}$ to $\bm{\basalfriction}_k$, plus one adjoint Stokes solve to compute the gradient at $\bm{\basalfriction}_k$, plus one incremental forward and one incremental adjoint Stokes solve per PCG iteration for solving the Newton system. Under PSF (5) and in row Iter 3, we write $6+22$ to indicate that $6$ Stokes solves were used during the standard course of the iteration, and $22$ Stokes solves were used to build the PSF (5) preconditioner.}
	\label{tab:newton_convergence_table}
\end{table}
